---
title: "RAW-C Data: Psychometric Predictive Power"
author: "Sean Trott"
date: "2-23-2026"
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
---

```{r, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "png")
```


# Setup

Load libraries

```{r setup}
library(tidyverse)
library(lme4)
library(broom.mixed)
library(lmerTest)
library(purrr)
library(ggrepel)
```


# Load data

**TODO**: Also compare to PPP for cosine distance over time.

```{r}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/Ambiguity/SSD/llm_sbe/src/analysis")

# df_pythia = read_csv("../../data/processed/rawc/pythia-14m_surprisals.csv")
### First, load all pythia models
directory_path <- "../../data/processed/rawc/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_pythia_models <- bind_rows(csv_list) %>%
  mutate(model_shorthand = sub("^EleutherAI/", "", model)) 
nrow(df_pythia_models)
table(df_pythia_models$model_shorthand)


### Load human dominance data
df_dominance = read_csv("../../data/human/dominance_norms_with_order.csv")


### Load human experiments
df_human_full_e1 = read_csv("../../data/human/polysemy_s1_final.csv") %>%
  select(Word, string, rt, log_rt, correct_response, version_with_order,
         subject, prior_rt, prior_accuracy) %>%
  mutate(experiment = "Experiment 1")
nrow(df_human_full_e1)
df_human_full_e2 = read_csv("../../data/human/polysemy_s2_final.csv") %>%
  select(Word, string, rt, log_rt, correct_response, version_with_order,
         subject, prior_rt, prior_accuracy) %>%
  mutate(experiment = "Experiment 2")
nrow(df_human_full_e2)

df_human_all = df_human_full_e2 %>%
  bind_rows(df_human_full_e1)
nrow(df_human_all)




### Merge data
## TODO: Check this
df_human_with_lm_full = df_pythia_models %>%
  # filter(checkpoint == 143000) %>%
  # filter(model_shorthand == "pythia-14m") %>%
  inner_join(df_human_all)


### Now human with dominance
df_human_with_lm_dominance = df_pythia_models %>%
  inner_join(df_dominance) %>%
  inner_join(df_human_all)


##### CENTER PREDICTORS
df_human_with_lm_full <- df_human_with_lm_full %>%
  mutate(
    ambiguity_c = if_else(ambiguity_type == "Polysemy", 0.5, -0.5),
    # dominance_z = scale(dominance_right)[,1],
    same_label = if_else(same == TRUE, "Same Sense", "Different Sense"),
    same_c = if_else(same == TRUE, 0.5, -0.5),
    surprisal_z = scale(target_surprisal_mean)
  )

df_human_with_lm_dominance <- df_human_with_lm_dominance %>%
  mutate(
    ambiguity_c = if_else(ambiguity_type == "Polysemy", 0.5, -0.5),
    dominance_z = scale(dominance_right)[,1],
    same_label = if_else(same == TRUE, "Same Sense", "Different Sense"),
    surprisal_z = scale(target_surprisal_mean)
  )


```


# Pre-registered analyses

Here, we run and report pre-registered analyses focusing on the ability of surprisal to account for experimental variables, i.e., sense dominance and the existence of a sense boundary.

In each case, we examine the impact on RT and also on accuracy.

## RQ2: Does surprisal account for dominance effect?


```{r}
df_human_with_lm_dominance_final_step = df_human_with_lm_dominance %>%
  filter(checkpoint == 143000)
df_human_with_lm_dominance_final_step_rt = df_human_with_lm_dominance_final_step %>%
  filter(correct_response == TRUE)




```

### RT Analysis

```{r dom_rt}
fit_dominance_test <- function(df_sub, key) {
  message("Fitting: ", key$model_shorthand, " (n = ", nrow(df_sub), ")  ",
          format(Sys.time(), "%H:%M:%S"))
  
  m_full <- lmer(log_rt ~ dominance_z + ambiguity_c + surprisal_z +
                   (1|subject) + (1|target) + (1|word) + (1|experiment),
                 data = df_sub, REML = FALSE)
  m_reduced <- lmer(log_rt ~ ambiguity_c + surprisal_z +
                      (1|subject) + (1|target) + (1|word) + (1|experiment),
                    data = df_sub, REML = FALSE)
  
  lrt <- anova(m_reduced, m_full)
  
  tibble(
    chisq    = lrt$Chisq[2],
    df       = lrt$Df[2],
    p_value  = lrt$`Pr(>Chisq)`[2],
    aic_full = AIC(m_full),
    aic_red  = AIC(m_reduced),
    dom_beta = fixef(m_full)[["dominance_z"]],
    dom_se   = sqrt(diag(vcov(m_full)))[["dominance_z"]],
    surprisal_beta = fixef(m_full)[["surprisal_z"]],
    surprisal_se   = sqrt(diag(vcov(m_full)))[["surprisal_z"]],
    ambiguity_beta = fixef(m_full)[["ambiguity_c"]],
    ambiguity_se   = sqrt(diag(vcov(m_full)))[["ambiguity_c"]]
  )
}

results <- df_human_with_lm_dominance_final_step_rt %>%
  group_by(model_shorthand) %>%
  group_modify(fit_dominance_test) %>%
  ungroup()

results

summary(results$chisq)
summary(results$dom_beta)
summary(results$surprisal_beta)
summary(results$ambiguity_beta)


df_pythia_models_params = df_pythia_models %>%
  distinct(model_shorthand, n_params)

results %>%
  inner_join(df_pythia_models_params) %>%
  ggplot(aes(x = n_params,
             y = chisq)) +
  geom_point(size = 6,
             alpha = .5) +
  scale_x_log10() +
  geom_text_repel(aes(label=model_shorthand), size=3) +
  labs(x = "Parameters",
       y = "LRT Chisq") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") 


results %>%
  inner_join(df_pythia_models_params) %>%
  ggplot(aes(x = n_params,
             y = surprisal_beta)) +
  geom_point(size = 6,
             alpha = .5) +
  scale_x_log10() +
  geom_text_repel(aes(label=model_shorthand), size=3) +
  labs(x = "Parameters",
       y = "Surprisal Coefficient (RT)") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") 


results = results %>%
  inner_join(df_pythia_models_params)

cor.test(log10(results$n_params), results$surprisal_beta)
cor.test(log10(results$n_params), results$chisq)

```


### Accuracy Analysis

```{r dom_accuracy}
fit_dominance_test_acc <- function(df_sub, key) {
  message("Fitting: ", key$model_shorthand, " (n = ", nrow(df_sub), ")  ",
          format(Sys.time(), "%H:%M:%S"))
  
  ctrl <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
  
  m_full <- glmer(correct_response ~ dominance_z + ambiguity_c + surprisal_z +
                    (1|subject) + (1|target) + (1|word) + (1|experiment),
                  data = df_sub, family = binomial, control = ctrl)
  m_reduced <- glmer(correct_response ~ ambiguity_c + surprisal_z +
                       (1|subject) + (1|target) + (1|word) + (1|experiment),
                     data = df_sub, family = binomial, control = ctrl)
  
  lrt <- anova(m_reduced, m_full)
  
  tibble(
    chisq    = lrt$Chisq[2],
    df       = lrt$Df[2],
    p_value  = lrt$`Pr(>Chisq)`[2],
    aic_full = AIC(m_full),
    aic_red  = AIC(m_reduced),
    dom_beta = fixef(m_full)[["dominance_z"]],
    dom_se   = sqrt(diag(vcov(m_full)))[["dominance_z"]],
    surprisal_beta = fixef(m_full)[["surprisal_z"]],
    surprisal_se   = sqrt(diag(vcov(m_full)))[["surprisal_z"]],
    ambiguity_beta = fixef(m_full)[["ambiguity_c"]],
    ambiguity_se   = sqrt(diag(vcov(m_full)))[["ambiguity_c"]],
    singular = isSingular(m_full)
  )
}

results_acc <- df_human_with_lm_dominance_final_step %>%
  group_by(model_shorthand) %>%
  group_modify(fit_dominance_test_acc) %>%
  ungroup()

results_acc
summary(results_acc$chisq)
summary(results_acc$dom_beta)
summary(results_acc$surprisal_beta)
summary(results_acc$ambiguity_beta)

results_acc <- results_acc %>%
  inner_join(df_pythia_models_params)

results_acc %>%
  ggplot(aes(x = n_params, y = chisq)) +
  geom_point(size = 6, alpha = .5) +
  scale_x_log10() +
  geom_text_repel(aes(label = model_shorthand), size = 3) +
  labs(x = "Parameters", y = "LRT Chisq (Accuracy)") +
  theme_minimal() +
  theme(text = element_text(size = 15), legend.position = "bottom")

results_acc %>%
  ggplot(aes(x = n_params, y = surprisal_beta)) +
  geom_point(size = 6, alpha = .5) +
  scale_x_log10() +
  geom_text_repel(aes(label = model_shorthand), size = 3) +
  labs(x = "Parameters", y = "Surprisal Coefficient (Accuracy)") +
  theme_minimal() +
  theme(text = element_text(size = 15), legend.position = "bottom")

cor.test(log10(results_acc$n_params), results_acc$surprisal_beta)
cor.test(log10(results_acc$n_params), results_acc$chisq)

```



### RQ6: Developmental trajectory

```{r dev_dom}
# ---- helper: aggregate + fit per (model, checkpoint) ----
# `group_var` is the non-surprisal covariate in the base model
# (e.g., "same" for sense boundary, "dominance_z" for dominance).
# If NULL, the base model is length-only.
fit_delta_r2 <- function(df, dv, group_var = NULL) {
  base_rhs <- if (is.null(group_var)) "length" else paste(group_var, "+ length")
  full_rhs <- paste(base_rhs, "+ mean_surprisal")
  
  # Group by the covariate too so the item-level aggregation keeps it
  group_cols <- c("model_shorthand", "n_params", "checkpoint",
                  group_var, "sentence1", "sentence2")
  
  df %>%
    mutate(length = stringr::str_count(sentence2, "\\S+")) %>%
    group_by(across(all_of(group_cols))) %>%
    summarise(
      dv_mean        = mean(.data[[dv]]),
      mean_surprisal = mean(surprisal_z),
      length         = first(length),
      .groups = "drop"
    ) %>%
    group_by(model_shorthand, n_params, checkpoint) %>%
    nest() %>%
    mutate(
      r2_base  = map_dbl(data, ~ summary(lm(as.formula(paste("dv_mean ~", base_rhs)), data = .x))$r.squared),
      r2_full  = map_dbl(data, ~ summary(lm(as.formula(paste("dv_mean ~", full_rhs)), data = .x))$r.squared),
      delta_r2 = r2_full - r2_base
    ) %>%
    ungroup() %>%
    select(model_shorthand, n_params, checkpoint, delta_r2)
}

# ---- helper: plot ΔR² developmental trajectory ----
plot_delta_r2 <- function(results, y_label) {
  df_summary <- results %>%
    group_by(checkpoint) %>%
    summarise(mean_delta = mean(delta_r2, na.rm = TRUE),
              se = sd(delta_r2) / sqrt(n()),
              .groups = "drop")
  
  ggplot(results,
         aes(x = checkpoint, y = delta_r2,
             color = reorder(model_shorthand, n_params))) +
    geom_line(alpha = 0.25) +
    geom_ribbon(data = df_summary,
                aes(x = checkpoint,
                    ymin = mean_delta - se,
                    ymax = mean_delta + se),
                inherit.aes = FALSE,
                alpha = 0.25, fill = "black") +
    geom_line(data = df_summary,
              aes(x = checkpoint, y = mean_delta),
              inherit.aes = FALSE,
              size = 1.5, color = "black") +
    scale_x_log10() +
    geom_vline(xintercept = 1000, linetype = "dotted", size = 1.2) +
    labs(x = "Training Step (Log10)", y = y_label, color = "") +
    theme_minimal() +
    theme(text = element_text(size = 15),
          legend.position = "bottom",
          legend.text = element_text(size = 11)) +
    guides(color = guide_legend(nrow = 2, byrow = TRUE)) +
    scale_color_manual(values = viridisLite::viridis(8, option = "magma",
                                                     begin = 0.8, end = 0.15))
}

# ---- fit all four ----
df_human_with_lm_dominance_rt <- df_human_with_lm_dominance %>%
  filter(correct_response == TRUE)

res_acc_with    <- fit_delta_r2(df_human_with_lm_dominance,    "correct_response", group_var = "dominance_z")
res_acc_without <- fit_delta_r2(df_human_with_lm_dominance,    "correct_response", group_var = NULL)
res_rt_with     <- fit_delta_r2(df_human_with_lm_dominance_rt, "log_rt",           group_var = "dominance_z")
res_rt_without  <- fit_delta_r2(df_human_with_lm_dominance_rt, "log_rt",           group_var = NULL)

# ---- four individual plots ----
plot_delta_r2(res_acc_with,    expression(Delta*R^2 ~ "(Acc., controlling dom. + length)"))
plot_delta_r2(res_acc_without, expression(Delta*R^2 ~ "(Acc., controlling length)"))
plot_delta_r2(res_rt_with,     expression(Delta*R^2 ~ "(RT, controlling dom. + length)"))
plot_delta_r2(res_rt_without,  expression(Delta*R^2 ~ "(RT, controlling length)"))

```



## RQ4: Does surprisal account for sense boundary effects?


## RQ2: Does surprisal account for dominance effect?


```{r}
df_human_with_lm_full_final_step = df_human_with_lm_full %>%
  filter(checkpoint == 143000)
df_human_with_lm_full_final_step_rt = df_human_with_lm_full_final_step %>%
  filter(correct_response == TRUE)


```

### RT Analysis

```{r sb_rt}
fit_sense_test <- function(df_sub, key) {
  message("Fitting: ", key$model_shorthand, " (n = ", nrow(df_sub), ")  ",
          format(Sys.time(), "%H:%M:%S"))
  
  m_full <- lmer(log_rt ~ same_c + ambiguity_c + surprisal_z +
                   (1|subject) + (1|target) + (1|word) + (1|experiment),
                 data = df_sub, REML = FALSE)
  m_reduced <- lmer(log_rt ~ ambiguity_c + surprisal_z +
                      (1|subject) + (1|target) + (1|word) + (1|experiment),
                    data = df_sub, REML = FALSE)
  
  lrt <- anova(m_reduced, m_full)
  
  tibble(
    chisq          = lrt$Chisq[2],
    df             = lrt$Df[2],
    p_value        = lrt$`Pr(>Chisq)`[2],
    aic_full       = AIC(m_full),
    aic_red        = AIC(m_reduced),
    same_beta      = fixef(m_full)[["same_c"]],
    same_se        = sqrt(diag(vcov(m_full)))[["same_c"]],
    surprisal_beta = fixef(m_full)[["surprisal_z"]],
    surprisal_se   = sqrt(diag(vcov(m_full)))[["surprisal_z"]],
    ambiguity_beta = fixef(m_full)[["ambiguity_c"]],
    ambiguity_se   = sqrt(diag(vcov(m_full)))[["ambiguity_c"]]
  )
}

results_sense_rt <- df_human_with_lm_full_final_step_rt %>%
  group_by(model_shorthand) %>%
  group_modify(fit_sense_test) %>%
  ungroup()

results_sense_rt
summary(results_sense_rt$chisq)
summary(results_sense_rt$same_beta)
summary(results_sense_rt$surprisal_beta)
summary(results_sense_rt$ambiguity_beta)

results_sense_rt <- results_sense_rt %>%
  inner_join(df_pythia_models_params)

results_sense_rt %>%
  ggplot(aes(x = n_params, y = chisq)) +
  geom_point(size = 6, alpha = .5) +
  scale_x_log10() +
  geom_text_repel(aes(label = model_shorthand), size = 3) +
  labs(x = "Parameters", y = "LRT Chisq (same_c)") +
  theme_minimal() +
  theme(text = element_text(size = 15), legend.position = "bottom")

results_sense_rt %>%
  ggplot(aes(x = n_params, y = surprisal_beta)) +
  geom_point(size = 6, alpha = .5) +
  scale_x_log10() +
  geom_text_repel(aes(label = model_shorthand), size = 3) +
  labs(x = "Parameters", y = "Surprisal Coefficient (RT)") +
  theme_minimal() +
  theme(text = element_text(size = 15), legend.position = "bottom")

cor.test(log10(results_sense_rt$n_params), results_sense_rt$surprisal_beta)
cor.test(log10(results_sense_rt$n_params), results_sense_rt$chisq)
```


### Accuracy analysis

```{r acc_comp_same}
fit_sense_test_acc <- function(df_sub, key) {
  message("Fitting: ", key$model_shorthand, " (n = ", nrow(df_sub), ")  ",
          format(Sys.time(), "%H:%M:%S"))
  
  ctrl <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
  
  m_full <- glmer(correct_response ~ same_c + ambiguity_c + surprisal_z +
                    (1|subject) + (1|target) + (1|word) + (1|experiment),
                  data = df_sub, family = binomial, control = ctrl)
  m_reduced <- glmer(correct_response ~ ambiguity_c + surprisal_z +
                       (1|subject) + (1|target) + (1|word) + (1|experiment),
                     data = df_sub, family = binomial, control = ctrl)
  
  lrt <- anova(m_reduced, m_full)
  
  tibble(
    chisq          = lrt$Chisq[2],
    df             = lrt$Df[2],
    p_value        = lrt$`Pr(>Chisq)`[2],
    aic_full       = AIC(m_full),
    aic_red        = AIC(m_reduced),
    same_beta      = fixef(m_full)[["same_c"]],
    same_se        = sqrt(diag(vcov(m_full)))[["same_c"]],
    surprisal_beta = fixef(m_full)[["surprisal_z"]],
    surprisal_se   = sqrt(diag(vcov(m_full)))[["surprisal_z"]],
    ambiguity_beta = fixef(m_full)[["ambiguity_c"]],
    ambiguity_se   = sqrt(diag(vcov(m_full)))[["ambiguity_c"]],
    singular       = isSingular(m_full)
  )
}

results_sense_acc <- df_human_with_lm_full_final_step %>%
  group_by(model_shorthand) %>%
  group_modify(fit_sense_test_acc) %>%
  ungroup()

results_sense_acc
summary(results_sense_acc$chisq)
summary(results_sense_acc$same_beta)
summary(results_sense_acc$surprisal_beta)
summary(results_sense_acc$ambiguity_beta)

results_sense_acc <- results_sense_acc %>%
  inner_join(df_pythia_models_params)

results_sense_acc %>%
  ggplot(aes(x = n_params, y = chisq)) +
  geom_point(size = 6, alpha = .5) +
  scale_x_log10() +
  geom_text_repel(aes(label = model_shorthand), size = 3) +
  labs(x = "Parameters", y = "LRT Chisq (same_c)") +
  theme_minimal() +
  theme(text = element_text(size = 15), legend.position = "bottom")

results_sense_acc %>%
  ggplot(aes(x = n_params, y = surprisal_beta)) +
  geom_point(size = 6, alpha = .5) +
  scale_x_log10() +
  geom_text_repel(aes(label = model_shorthand), size = 3) +
  labs(x = "Parameters", y = "Surprisal Coefficient (Accuracy)") +
  theme_minimal() +
  theme(text = element_text(size = 15), legend.position = "bottom")

cor.test(log10(results_sense_acc$n_params), results_sense_acc$surprisal_beta)
cor.test(log10(results_sense_acc$n_params), results_sense_acc$chisq)
```



### RQ6: Developmental trajectory

```{r dev_same}
# ---- helper: aggregate + fit per (model, checkpoint) ----
fit_delta_r2 <- function(df, dv, include_same) {
  base_rhs <- if (include_same) "same + length" else "length"
  full_rhs <- paste(base_rhs, "+ mean_surprisal")
  
  df %>%
    mutate(length = stringr::str_count(sentence2, "\\S+")) %>%
    group_by(model_shorthand, n_params, checkpoint, same, sentence1, sentence2) %>%
    summarise(
      dv_mean        = mean(.data[[dv]]),
      mean_surprisal = mean(surprisal_z),
      length         = first(length),
      .groups = "drop"
    ) %>%
    group_by(model_shorthand, n_params, checkpoint) %>%
    nest() %>%
    mutate(
      r2_base  = map_dbl(data, ~ summary(lm(as.formula(paste("dv_mean ~", base_rhs)), data = .x))$r.squared),
      r2_full  = map_dbl(data, ~ summary(lm(as.formula(paste("dv_mean ~", full_rhs)), data = .x))$r.squared),
      delta_r2 = r2_full - r2_base
    ) %>%
    ungroup() %>%
    select(model_shorthand, n_params, checkpoint, delta_r2)
}

# ---- helper: plot ΔR² developmental trajectory ----
plot_delta_r2 <- function(results, y_label) {
  df_summary <- results %>%
    group_by(checkpoint) %>%
    summarise(mean_delta = mean(delta_r2, na.rm = TRUE),
              se = sd(delta_r2) / sqrt(n()),
              .groups = "drop")
  
  ggplot(results,
         aes(x = checkpoint, y = delta_r2,
             color = reorder(model_shorthand, n_params))) +
    geom_line(alpha = 0.25) +
    geom_ribbon(data = df_summary,
                aes(x = checkpoint,
                    ymin = mean_delta - se,
                    ymax = mean_delta + se),
                inherit.aes = FALSE,
                alpha = 0.25, fill = "black") +
    geom_line(data = df_summary,
              aes(x = checkpoint, y = mean_delta),
              inherit.aes = FALSE,
              size = 1.5, color = "black") +
    scale_x_log10() +
    geom_vline(xintercept = 1000, linetype = "dotted", size = 1.2) +
    labs(x = "Training Step (Log10)", y = y_label, color = "") +
    theme_minimal() +
    theme(text = element_text(size = 15),
          legend.position = "bottom",
          legend.text = element_text(size = 11)) +
    guides(color = guide_legend(nrow = 2, byrow = TRUE)) +
    scale_color_manual(values = viridisLite::viridis(8, option = "magma",
                                                     begin = 0.8, end = 0.15))
}

# ---- fit all four ----
df_human_with_lm_full_rt <- df_human_with_lm_full %>% filter(correct_response == TRUE)

res_acc_with    <- fit_delta_r2(df_human_with_lm_full,    "correct_response", include_same = TRUE)
res_acc_without <- fit_delta_r2(df_human_with_lm_full,    "correct_response", include_same = FALSE)
res_rt_with     <- fit_delta_r2(df_human_with_lm_full_rt, "log_rt",           include_same = TRUE)
res_rt_without  <- fit_delta_r2(df_human_with_lm_full_rt, "log_rt",           include_same = FALSE)

# ---- four individual plots ----
plot_delta_r2(res_acc_with,    expression(Delta*R^2 ~ "(Acc., controlling same + length)"))
plot_delta_r2(res_acc_without, expression(Delta*R^2 ~ "(Acc., controlling length"))
plot_delta_r2(res_rt_with,     expression(Delta*R^2 ~ "(RT, controlling same + length)"))
plot_delta_r2(res_rt_without,  expression(Delta*R^2 ~ "(RT, controlling length)"))
```


