---
title: "RAW-C Data: Psychometric Predictive Power"
author: "Sean Trott"
date: "2-23-2026"
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
---

```{r, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "png")
```


# Setup

Load libraries

```{r setup}
library(tidyverse)
library(lme4)
library(broom.mixed)
library(lmerTest)
library(furrr)
```


# Load data

**TODO**: Also compare to PPP for cosine distance over time.

```{r}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/Ambiguity/SSD/llm_sbe/src/analysis")

# df_pythia = read_csv("../../data/processed/rawc/pythia-14m_surprisals.csv")
### First, load all pythia models
directory_path <- "../../data/processed/rawc/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_pythia_models <- bind_rows(csv_list) %>%
  mutate(model_shorthand = sub("^EleutherAI/", "", model)) 
nrow(df_pythia_models)
table(df_pythia_models$model_shorthand)


### Load human dominance data
df_dominance = read_csv("../../data/human/dominance_norms_with_order.csv")


### Load human experiments
df_human_full_e1 = read_csv("../../data/human/polysemy_s1_final.csv") %>%
  select(Word, string, rt, log_rt, correct_response, version_with_order,
         subject, prior_rt, prior_accuracy) %>%
  mutate(experiment = "Experiment 1")
nrow(df_human_full_e1)
df_human_full_e2 = read_csv("../../data/human/polysemy_s2_final.csv") %>%
  select(Word, string, rt, log_rt, correct_response, version_with_order,
         subject, prior_rt, prior_accuracy) %>%
  mutate(experiment = "Experiment 2")
nrow(df_human_full_e2)

df_human_all = df_human_full_e2 %>%
  bind_rows(df_human_full_e1)
nrow(df_human_all)




### Merge data
## TODO: Check this
df_human_with_lm_full = df_pythia_models %>%
  # filter(checkpoint == 143000) %>%
  inner_join(df_human_all)


### Now human with dominance
df_human_with_lm_dominance = df_pythia_models %>%
  inner_join(df_dominance) %>%
  inner_join(df_human_all)


##### CENTER PREDICTORS
df_human_with_lm_full <- df_human_with_lm_full %>%
  mutate(
    ambiguity_c = if_else(ambiguity_type == "Polysemy", 0.5, -0.5),
    # dominance_z = scale(dominance_right)[,1],
    same_label = if_else(same == TRUE, "Same Sense", "Different Sense"),
    same_c = if_else(same == TRUE, 0.5, -0.5),
    surprisal_z = scale(target_surprisal_mean)
  )

df_human_with_lm_dominance <- df_human_with_lm_dominance %>%
  mutate(
    ambiguity_c = if_else(ambiguity_type == "Polysemy", 0.5, -0.5),
    dominance_z = scale(dominance_right)[,1],
    same_label = if_else(same == TRUE, "Same Sense", "Different Sense"),
    surprisal_z = scale(target_surprisal_mean)
  )


```



# Pre-registered analyses

Here, we run and report pre-registered analyses focusing on the ability of surprisal to account for experimental variables, i.e., sense dominance and the existence of a sense boundary.

In each case, we examine the impact on RT and also on accuracy.

## RQ2: Does surprisal account for dominance effect?


```{r}

plan(multisession, workers = 7)

# --- Prep data once ---
df_prepped <- df_human_with_lm_dominance %>%
  mutate(
    ambiguity_c = if_else(ambiguity_type == "Polysemy", 0.5, -0.5),
    dominance_z = scale(dominance_right)[,1]
  ) %>%
  group_by(model, model_shorthand, n_params, checkpoint) %>%
  mutate(surprisal_z = scale(target_surprisal_mean)[,1]) %>%
  ungroup()

# --- Helper: fit one group ---
fit_rq2_rt <- function(.x) {
  meta <- .x %>% distinct(model, model_shorthand, n_params, checkpoint) %>% slice(1)
  
  full <- tryCatch(
    lmer(log_rt ~ dominance_z + ambiguity_c + surprisal_z +
           (1|subject) + (1|target) + (1|word) + (1|experiment),
         data = .x, REML = FALSE),
    error = function(e) NULL)
  reduced <- tryCatch(
    lmer(log_rt ~ ambiguity_c + surprisal_z +
           (1|subject) + (1|target) + (1|word) + (1|experiment),
         data = .x, REML = FALSE),
    error = function(e) NULL)
  lrt <- if (!is.null(full) && !is.null(reduced)) anova(full, reduced) else NULL
  
  tibble(
    model            = meta$model,
    model_shorthand  = meta$model_shorthand,
    n_params         = meta$n_params,
    checkpoint       = meta$checkpoint,
    rq2_rt_dom_beta    = if (!is.null(full)) fixef(full)["dominance_z"] else NA_real_,
    rq2_rt_dom_se      = if (!is.null(full)) sqrt(diag(vcov(full)))["dominance_z"] else NA_real_,
    rq2_rt_surp_beta   = if (!is.null(full)) fixef(full)["surprisal_z"] else NA_real_,
    rq2_rt_surp_se     = if (!is.null(full)) sqrt(diag(vcov(full)))["surprisal_z"] else NA_real_,
    rq2_rt_lrt_chi2    = if (!is.null(lrt)) lrt$Chisq[2] else NA_real_,
    rq2_rt_lrt_p       = if (!is.null(lrt)) lrt$`Pr(>Chisq)`[2] else NA_real_,
    rq2_rt_aic_full    = if (!is.null(full)) AIC(full) else NA_real_,
    rq2_rt_aic_reduced = if (!is.null(reduced)) AIC(reduced) else NA_real_,
    rq2_rt_aic_delta   = rq2_rt_aic_reduced - rq2_rt_aic_full
  )
}

fit_rq2_acc <- function(.x) {
  meta <- .x %>% distinct(model, model_shorthand, n_params, checkpoint) %>% slice(1)
  
  full <- tryCatch(
    glmer(correct_response ~ dominance_z + ambiguity_c + surprisal_z +
            (1|subject) + (1|target) + (1|word) + (1|experiment),
          data = .x, family = binomial, nAGQ = 0,
          control = glmerControl(optimizer = "bobyqa")),
    error = function(e) NULL)
  reduced <- tryCatch(
    glmer(correct_response ~ ambiguity_c + surprisal_z +
            (1|subject) + (1|target) + (1|word) + (1|experiment),
          data = .x, family = binomial, nAGQ = 0,
          control = glmerControl(optimizer = "bobyqa")),
    error = function(e) NULL)
  lrt <- if (!is.null(full) && !is.null(reduced)) anova(full, reduced) else NULL
  
  tibble(
    model            = meta$model,
    model_shorthand  = meta$model_shorthand,
    n_params         = meta$n_params,
    checkpoint       = meta$checkpoint,
    rq2_acc_dom_beta    = if (!is.null(full)) fixef(full)["dominance_z"] else NA_real_,
    rq2_acc_dom_se      = if (!is.null(full)) sqrt(diag(vcov(full)))["dominance_z"] else NA_real_,
    rq2_acc_surp_beta   = if (!is.null(full)) fixef(full)["surprisal_z"] else NA_real_,
    rq2_acc_surp_se     = if (!is.null(full)) sqrt(diag(vcov(full)))["surprisal_z"] else NA_real_,
    rq2_acc_lrt_chi2    = if (!is.null(lrt)) lrt$Chisq[2] else NA_real_,
    rq2_acc_lrt_p       = if (!is.null(lrt)) lrt$`Pr(>Chisq)`[2] else NA_real_,
    rq2_acc_aic_full    = if (!is.null(full)) AIC(full) else NA_real_,
    rq2_acc_aic_reduced = if (!is.null(reduced)) AIC(reduced) else NA_real_,
    rq2_acc_aic_delta   = rq2_acc_aic_reduced - rq2_acc_aic_full
  )
}

# --- Helper: run with checkpoint saving ---
run_with_checkpoint <- function(groups, fit_fn, save_path) {
  # Load previous progress
  completed <- if (file.exists(save_path)) readRDS(save_path) else tibble()
  
  if (nrow(completed) > 0) {
    message(sprintf("Loaded %d completed fits from %s", nrow(completed), save_path))
  }
  
  # Filter to remaining groups
  remaining <- keep(groups, function(.x) {
    meta <- .x %>% distinct(model_shorthand, checkpoint) %>% slice(1)
    !any(completed$model_shorthand == meta$model_shorthand & 
         completed$checkpoint == meta$checkpoint)
  })
  
  message(sprintf("%d groups remaining out of %d total", length(remaining), length(groups)))
  
  if (length(remaining) == 0) {
    message("All groups already completed!")
    return(completed)
  }
  
  # Run in batches so we can save periodically
  batch_size <- 7  # save after every batch
  batches <- split(remaining, ceiling(seq_along(remaining) / batch_size))
  
  for (i in seq_along(batches)) {
    message(sprintf("Batch %d/%d (%d fits)...", i, length(batches), length(batches[[i]])))
    
    batch_results <- future_map(batches[[i]], fit_fn, .progress = TRUE) %>% bind_rows()
    completed <- bind_rows(completed, batch_results)
    saveRDS(completed, save_path)
    
    message(sprintf("  Saved. %d/%d total complete.", nrow(completed), length(groups)))
  }
  
  completed %>% arrange(n_params, checkpoint)
}

# --- Split groups ---
rt_groups <- df_prepped %>%
  filter(correct_response == TRUE) %>%
  group_by(model, model_shorthand, n_params, checkpoint) %>%
  group_split()

acc_groups <- df_prepped %>%
  mutate(correct_response = as.numeric(correct_response)) %>%
  group_by(model, model_shorthand, n_params, checkpoint) %>%
  group_split()

# --- Run! ---
message("=== Running RQ2 RT models ===")
developmental_results_rq2_rt <- run_with_checkpoint(
  rt_groups, fit_rq2_rt, "rq2_rt_checkpoint.rds"
)

message("=== Running RQ2 Accuracy models ===")
developmental_results_rq2_acc <- run_with_checkpoint(
  acc_groups, fit_rq2_acc, "rq2_acc_checkpoint.rds"
)

# --- Save final results ---
saveRDS(developmental_results_rq2_rt, "developmental_results_rq2_rt.rds")
saveRDS(developmental_results_rq2_acc, "developmental_results_rq2_acc.rds")




```




### RQ6: Developmental trajectory

```{r}
developmental_results_rq2_rt %>%
  ggplot(aes(x = checkpoint,
             y = rq2_rt_surp_beta,
             color = reorder(model_shorthand, n_params))) +
  geom_point(size = 4, alpha = .7) +
  geom_line(size = 1.3) +
  geom_point(size = 4, alpha = .4) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Coefficient (Surprisal)",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  scale_shape_manual(values = c("FALSE" = 1, "TRUE" = 16),
                     labels = c("n.s.", "p < .05"),
                     name = NULL) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(4, option = "mako", 
                                                begin = 0.8, end = 0.15))

developmental_results_rq2_rt %>%
  ggplot(aes(x = checkpoint,
             y = rq2_rt_aic_delta,
             color = reorder(model_shorthand, n_params))) +
  geom_point(size = 4, alpha = .7) +
  geom_line(size = 1.3) +
  geom_point(size = 4, alpha = .4) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "AIC Delta (with vs. without Dominance)",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  scale_shape_manual(values = c("FALSE" = 1, "TRUE" = 16),
                     labels = c("n.s.", "p < .05"),
                     name = NULL) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(4, option = "mako", 
                                                begin = 0.8, end = 0.15))

developmental_results_rq2_rt %>%
  ggplot(aes(x = checkpoint,
             y = rq2_rt_lrt_chi2,
             color = reorder(model_shorthand, n_params))) +
  geom_point(size = 4, alpha = .7) +
  geom_line(size = 1.3) +
  geom_point(size = 4, alpha = .4) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "X^2 (with vs. without Dominance)",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  scale_shape_manual(values = c("FALSE" = 1, "TRUE" = 16),
                     labels = c("n.s.", "p < .05"),
                     name = NULL) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(4, option = "mako", 
                                                begin = 0.8, end = 0.15))
```




```{r}
# --- RT models (correct responses only) ---
developmental_results_rq2_rt <- df_human_with_lm_dominance %>%
  filter(correct_response == TRUE) %>%
  mutate(
    ambiguity_c = if_else(ambiguity_type == "Polysemy", 0.5, -0.5),
    dominance_z = scale(dominance_right)[,1],
    surprisal_z = scale(target_surprisal_mean)[,1]
  ) %>%
  group_by(model, model_shorthand, n_params, checkpoint) %>%
  group_map(~ {
    full <- tryCatch(
      lmer(log_rt ~ dominance_z + ambiguity_c + surprisal_z +
             (1|subject) + (1|target) + (1|word) + (1|experiment),
           data = .x, REML = FALSE),
      error = function(e) NULL)
    reduced <- tryCatch(
      lmer(log_rt ~ ambiguity_c + surprisal_z +
             (1|subject) + (1|target) + (1|word) + (1|experiment),
           data = .x, REML = FALSE),
      error = function(e) NULL)
    lrt <- if (!is.null(full) && !is.null(reduced)) anova(full, reduced) else NULL

    tibble(
      model            = .y$model,
      model_shorthand  = .y$model_shorthand,
      n_params         = .y$n_params,
      checkpoint       = .y$checkpoint,
      rq2_rt_dom_beta    = if (!is.null(full)) fixef(full)["dominance_z"] else NA_real_,
      rq2_rt_dom_se      = if (!is.null(full)) sqrt(diag(vcov(full)))["dominance_z"] else NA_real_,
      rq2_rt_surp_beta   = if (!is.null(full)) fixef(full)["surprisal_z"] else NA_real_,
      rq2_rt_surp_se     = if (!is.null(full)) sqrt(diag(vcov(full)))["surprisal_z"] else NA_real_,
      rq2_rt_lrt_chi2    = if (!is.null(lrt)) lrt$Chisq[2] else NA_real_,
      rq2_rt_lrt_p       = if (!is.null(lrt)) lrt$`Pr(>Chisq)`[2] else NA_real_,
      rq2_rt_aic_full    = if (!is.null(full)) AIC(full) else NA_real_,
      rq2_rt_aic_reduced = if (!is.null(reduced)) AIC(reduced) else NA_real_,
      rq2_rt_aic_delta   = rq2_rt_aic_reduced - rq2_rt_aic_full
    )
  }) %>%
  bind_rows() %>%
  arrange(n_params, checkpoint)



### TODO: Check below

# --- Accuracy models (all responses) ---
developmental_results_rq2_acc <- df_human_with_lm_dominance %>%
  mutate(
    ambiguity_c = if_else(ambiguity_type == "Polysemy", 0.5, -0.5),
    dominance_z = scale(dominance_right)[,1],
    surprisal_z = scale(surprisal)[,1],
    correct_response = as.numeric(correct_response)
  ) %>%
  group_by(model, model_shorthand, n_params, checkpoint) %>%
  group_map(~ {
    full <- tryCatch(
      glmer(correct_response ~ dominance_z + ambiguity_c + surprisal_z +
              (1|subject) + (1|target) + (1|word) + (1|experiment),
            data = .x, family = binomial),
      error = function(e) NULL)
    reduced <- tryCatch(
      glmer(correct_response ~ ambiguity_c + surprisal_z +
              (1|subject) + (1|target) + (1|word) + (1|experiment),
            data = .x, family = binomial),
      error = function(e) NULL)
    lrt <- if (!is.null(full) && !is.null(reduced)) anova(full, reduced) else NULL

    tibble(
      model            = .y$model,
      model_shorthand  = .y$model_shorthand,
      n_params         = .y$n_params,
      checkpoint       = .y$checkpoint,
      rq2_acc_dom_beta    = if (!is.null(full)) fixef(full)["dominance_z"] else NA_real_,
      rq2_acc_dom_se      = if (!is.null(full)) sqrt(diag(vcov(full)))["dominance_z"] else NA_real_,
      rq2_acc_surp_beta   = if (!is.null(full)) fixef(full)["surprisal_z"] else NA_real_,
      rq2_acc_surp_se     = if (!is.null(full)) sqrt(diag(vcov(full)))["surprisal_z"] else NA_real_,
      rq2_acc_lrt_chi2    = if (!is.null(lrt)) lrt$Chisq[2] else NA_real_,
      rq2_acc_lrt_p       = if (!is.null(lrt)) lrt$`Pr(>Chisq)`[2] else NA_real_,
      rq2_acc_aic_full    = if (!is.null(full)) AIC(full) else NA_real_,
      rq2_acc_aic_reduced = if (!is.null(reduced)) AIC(reduced) else NA_real_,
      rq2_acc_aic_delta   = rq2_acc_aic_reduced - rq2_acc_aic_full
    )
  }) %>%
  bind_rows() %>%
  arrange(n_params, checkpoint)
```



## RQ4: Does surprisal account for the sense boundary effect?






# Analyses

## RQs 1-2: Does surprisal track sense dominance?

### RQ 1

```{r}
### Dominance analysis
df_merged = df_pythia %>%
  inner_join(df_dominance)

df_merged = df_merged %>%
  mutate(Target = case_when(
    dominance_right > .5 ~ "Dominant",
    dominance_right < -.5 ~ "Subordinate",
    dominance_right >= -.5 & dominance_right <= .5 ~ "Neutral"
  ))


### How does surprisal track dominance of target
mod_full = lmer(data = df_merged,
             final_token_surprisal ~ ambiguity_type + dominance_right +
             (1 | word),
             REML = FALSE)
mod_reduced = lmer(data = df_merged,
             final_token_surprisal ~ ambiguity_type + # dominance_right +
             (1 | word),
             REML = FALSE)
anova(mod_full, mod_reduced)
```

### RQ 2

Does surprisal account for dominance effect in humans?


**TODO**: Maybe store human analyses in a separate .Rmd for ease.

```{r}
df_merged_all_human = df_human_all %>%
  inner_join(df_merged)


mod_rt_full = lmer(log_rt ~ dominance_right + ambiguity_type + final_token_surprisal + 
                (1|subject) + (1|target) + (1|word) + (1|experiment),
              data = df_merged_all_human %>% filter(correct_response == TRUE),
              REML = FALSE)

mod_rt_reduced = lmer(log_rt ~ ambiguity_type + final_token_surprisal + 
                (1|subject) + (1|target) + (1|word) + (1|experiment),
              data = df_merged_all_human %>% filter(correct_response == TRUE),
              REML = FALSE)

anova(mod_rt_full, mod_rt_reduced)
summary(mod_rt_full)



mod_accuracy_full = glmer(correct_response ~ dominance_right + 
                            ambiguity_type + 
                            final_token_surprisal + 
                (1|subject) + (1|target) + (1|word) + (1|experiment),
              data = df_merged_all_human,
              family = binomial())


mod_accuracy_reduced = glmer(correct_response ~ # dominance_right + 
                            ambiguity_type + 
                            final_token_surprisal + 
                (1|subject) + (1|target) + (1|word) + (1|experiment),
              data = df_merged_all_human,
              family = binomial())

anova(mod_accuracy_full, mod_accuracy_reduced)
summary(mod_accuracy_full)

```





```{r}
df_pythia %>%
  group_by(same, ambiguity_type) %>%
  summarise(
    mean_surprisal = mean(target_surprisal_mean, na.rm = TRUE),
    se_surprisal = sd(target_surprisal_mean, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = same,
             y = mean_surprisal,
             color = ambiguity_type)) +
  geom_pointrange(aes(ymin = mean_surprisal - se_surprisal,
                      ymax = mean_surprisal + se_surprisal),
                  position = position_dodge(width = 0.2),
                  size = 0.5) +
  labs(x = "Same meaning",
       y = "Target sentence mean surprisal",
       color = "Ambiguity type") +
  theme_minimal()

df_pythia %>%
  group_by(same, ambiguity_type) %>%
  summarise(
    mean_surprisal = mean(final_token_surprisal, na.rm = TRUE),
    se_surprisal = sd(final_token_surprisal, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = same,
             y = mean_surprisal,
             color = ambiguity_type)) +
  geom_pointrange(aes(ymin = mean_surprisal - se_surprisal,
                      ymax = mean_surprisal + se_surprisal),
                  position = position_dodge(width = 0.2),
                  size = 0.5) +
  labs(x = "Same meaning",
       y = "Final token surprisal",
       color = "Ambiguity type") +
  theme_minimal()




### Dominance analysis
df_merged = df_pythia %>%
  inner_join(df_dominance)

df_merged = df_merged %>%
  mutate(Target = case_when(
    dominance_right > .5 ~ "Dominant",
    dominance_right < -.5 ~ "Subordinate",
    dominance_right >= -.5 & dominance_right <= .5 ~ "Neutral"
  ))




df_merged %>%
  ggplot(aes(x = Target,
             y = final_token_surprisal,
             color = ambiguity_type)) +
  stat_summary (fun = function(x){mean(x)},
                fun.min = function(x){mean(x) - 2*sd(x)/sqrt(length(x))},
                fun.max = function(x){mean(x) + 2*sd(x)/sqrt(length(x))},
                geom= 'pointrange', 
                position=position_dodge(width=0.95)) +
  labs(x = "Target Dominance",
       y = "Target sentence mean surprisal",
       color = "Ambiguity Type") +
  theme_minimal() +
  facet_wrap(~Class)



### How does surprisal track dominance of target
summary(lmer(data = df_merged,
             final_token_surprisal ~ ambiguity_type + dominance_right +
               (1 | word)))



df_merged_human = df_human_e1 %>%
  inner_join(df_merged)


summary(lmer(data = df_merged_human,
             correct_response ~ ambiguity_type * dominance_right + target_surprisal_mean +
               (1 | word) +
               (1 | target)))


summary(lmer(data = df_merged_human,
             log_rt ~ ambiguity_type * dominance_right + target_surprisal_mean +
               (1 | word) +
               (1 | target)))


df_merged_human %>%
  ggplot(aes(x = target_surprisal_mean,
             y = log_rt)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()

df_merged_human %>%
  ggplot(aes(x = dominance_right,
             y = log_rt)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_minimal()

df <- df_merged_human %>%
  mutate(
    ambiguity_c = if_else(ambiguity_type == "Polysemy", 0.5, -0.5),
    dominance_z = scale(dominance_right)[,1],
    surprisal_z = scale(target_surprisal_mean)[,1]
  )

# --- Fit models ---

# Without surprisal
m_rt_no_surp  <- lmer(log_rt ~ ambiguity_c * dominance_z + (1|word) + (1|target), data = df)
m_acc_no_surp <- lmer(correct_response ~ ambiguity_c * dominance_z + (1|word) + (1|target), data = df)

# With surprisal
m_rt_surp  <- lmer(log_rt ~ ambiguity_c * dominance_z + surprisal_z + (1|word) + (1|target), data = df)
m_acc_surp <- lmer(correct_response ~ ambiguity_c * dominance_z + surprisal_z + (1|word) + (1|target), data = df)

# --- Extract and combine ---
all_coefs <- bind_rows(
  tidy(m_rt_no_surp, conf.int = TRUE)  %>% mutate(model = "Without surprisal", dv = "log RT"),
  tidy(m_rt_surp, conf.int = TRUE)     %>% mutate(model = "With surprisal",    dv = "log RT"),
  tidy(m_acc_no_surp, conf.int = TRUE) %>% mutate(model = "Without surprisal", dv = "Accuracy"),
  tidy(m_acc_surp, conf.int = TRUE)    %>% mutate(model = "With surprisal",    dv = "Accuracy")
) %>%
  filter(effect == "fixed", term != "(Intercept)")

# --- Plot ---
all_coefs %>%
  ggplot(aes(x = term, y = estimate, color = model)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high),
                  position = position_dodge(width = 0.3)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ dv, scales = "free_x") +
  labs(y = "Estimate (standardized)", x = NULL, color = NULL) +
  coord_flip() +
  theme_minimal()




#### Extract sense-level dominance
df_merged = df_merged %>%
  mutate(
    # Target is the second element: M1_a_M2_b → M2, M2_b_M1_a → M1
    target_meaning = str_extract(version_with_order, "M[12](?=_[ab]$)")
  )

sense_surprisal <- df_merged %>%
  group_by(word, target_meaning) %>%
  summarise(
    mean_surprisal = mean(target_surprisal_mean, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = target_meaning,
              values_from = mean_surprisal,
              names_prefix = "surprisal_") %>%
  mutate(
    surprisal_diff = surprisal_M2 - surprisal_M1
  )

sense_dominance <- df_merged %>%
  group_by(word, target_meaning) %>%
  summarise(
    mean_dominance = mean(dominance_right, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = target_meaning,
              values_from = mean_dominance,
              names_prefix = "dominance_") %>%
  mutate(
    dominance_diff = dominance_M2 - dominance_M1
  )

# Join and correlate
sense_level <- sense_surprisal %>%
  inner_join(sense_dominance, by = "word")

cor.test(sense_level$dominance_diff, sense_level$surprisal_diff)

# Visualize
sense_level %>%
  ggplot(aes(x = dominance_diff, y = surprisal_diff)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  labs(x = "Human dominance (M2 - M1)",
       y = "Surprisal (M2 - M1)") +
  theme_minimal()


### all human
df_merged_all_human = df_human_e1 %>%
  inner_join(df_pythia) %>%
  mutate(surprisal_z = scale(target_surprisal_mean, scale = FALSE))

summary(lmer(data = df_merged_all_human,
             log_rt ~ same + 
               ambiguity_type +
               surprisal_z +
               (1 | word) +
               (1 | target)))


summary(lmer(data = df_merged_all_human,
             correct_response ~ same + 
               ambiguity_type +
               target_surprisal_mean +
               (1 | word) +
               (1 | target)))


### Plot effect size
# Standardize continuous predictors
# Standardize predictors
df <- df_merged_all_human %>%
  mutate(
    same_c = if_else(same, 0.5, -0.5),
    ambiguity_c = if_else(ambiguity_type == "Polysemy", 0.5, -0.5),
    surprisal_z = scale(target_surprisal_mean)[,1]
  )

# --- Fit models ---

# RT
m_rt_no_surp <- lmer(log_rt ~ same_c + ambiguity_c + (1|word) + (1|target), data = df)
m_rt_surp    <- lmer(log_rt ~ same_c + ambiguity_c + surprisal_z + (1|word) + (1|target), data = df)

# Accuracy
m_acc_no_surp <- lmer(correct_response ~ same_c + ambiguity_c + (1|word) + (1|target), data = df)
m_acc_surp    <- lmer(correct_response ~ same_c + ambiguity_c + surprisal_z + (1|word) + (1|target), data = df)

# --- Extract and combine ---
all_coefs <- bind_rows(
  tidy(m_rt_no_surp, conf.int = TRUE)  %>% mutate(model = "Without surprisal", dv = "log RT"),
  tidy(m_rt_surp, conf.int = TRUE)     %>% mutate(model = "With surprisal",    dv = "log RT"),
  tidy(m_acc_no_surp, conf.int = TRUE) %>% mutate(model = "Without surprisal", dv = "Accuracy"),
  tidy(m_acc_surp, conf.int = TRUE)    %>% mutate(model = "With surprisal",    dv = "Accuracy")
) %>%
  filter(effect == "fixed", term != "(Intercept)")

# --- Plot ---
all_coefs %>%
  ggplot(aes(x = term, y = estimate, color = model)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high),
                  position = position_dodge(width = 0.3)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ dv, scales = "free_x") +
  labs(y = "Estimate (standardized)", x = NULL, color = NULL) +
  coord_flip() +
  theme_minimal()





### Full human data
df_merged_all_human = df_human_full %>%
  inner_join(df_pythia)



### TO DO: Decide whether to include prior RT/accuracy?
summary(lmer(data = df_merged_all_human,
             log_rt ~ same + ambiguity_type +
               target_surprisal_mean +
               (1 | word) +
               (1 | subject) +
               (1 | target)))


summary(glmer(data = df_merged_all_human,
             correct_response ~ same + 
               ambiguity_type +
               # prior_accuracy +
               target_surprisal_mean +
               (1 | word) +
               (1 | subject) +
               (1 | target),
             family = binomial()))


```



