---
title: "RAW-C Data: LM Analysis"
author: "Sean Trott"
date: "2-23-2026"
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
---

```{r, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "png")
```


# Setup

Load libraries

```{r setup}
library(tidyverse)
library(lme4)
library(broom.mixed)
library(lmerTest)
library(kableExtra)
library(ggridges)
library(ggrepel)
```


# Load data

```{r}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/Ambiguity/SSD/llm_sbe/src/analysis")

# df_pythia = read_csv("../../data/processed/rawc/pythia-14m_surprisals.csv")
### First, load all pythia models
directory_path <- "../../data/processed/rawc/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_pythia_models <- bind_rows(csv_list) %>%
  mutate(model_shorthand = sub("^EleutherAI/", "", model)) 
nrow(df_pythia_models)
table(df_pythia_models$model_shorthand)

### Load human dominance data
df_dominance = read_csv("../../data/human/dominance_norms_with_order.csv")


### Merge data
df_merged = df_pythia_models %>%
  inner_join(df_dominance)



##### CENTER PREDICTORS
df_merged <- df_merged %>%
  mutate(
    ambiguity_c = if_else(ambiguity_type == "Polysemy", 0.5, -0.5),
    dominance_z = scale(dominance_right)[,1]
  )

df_pythia_models <- df_pythia_models %>%
  mutate(
    ambiguity_c = if_else(ambiguity_type == "Polysemy", 0.5, -0.5),
    same_c = if_else(same == TRUE, 0.5, -0.5),
    same_label = if_else(same == TRUE, "Same Sense", "Different Sense")
  )
```





# Pre-registered analyses

*Note that analyses of human data (RQ2, RQ4) are included in a separate file.

## RQ1: Does surprisal of the TARGET phrase track sense dominance?


```{r}
developmental_results_rq1 <- df_merged %>%
  mutate(
    ambiguity_c = if_else(ambiguity_type == "Polysemy", 0.5, -0.5),
    dominance_z = scale(dominance_right)[,1]
  ) %>%
  group_by(model, model_shorthand, n_params, checkpoint) %>%
  group_map(~ {
    rq1_full    <- tryCatch(lmer(target_surprisal_mean ~ ambiguity_c + dominance_z + (1 | word), data = .x, REML = FALSE), error = function(e) NULL)
    rq1_reduced <- tryCatch(lmer(target_surprisal_mean ~ ambiguity_c +               (1 | word), data = .x, REML = FALSE), error = function(e) NULL)
    lrt <- if (!is.null(rq1_full) && !is.null(rq1_reduced)) anova(rq1_full, rq1_reduced) else NULL
    tibble(
      model            = .y$model,
      model_shorthand  = .y$model_shorthand,
      n_params         = .y$n_params,
      checkpoint       = .y$checkpoint,
      rq1_dom_beta     = if (!is.null(rq1_full)) fixef(rq1_full)["dominance_z"] else NA_real_,
      rq1_dom_se       = if (!is.null(rq1_full)) sqrt(diag(vcov(rq1_full)))["dominance_z"] else NA_real_,
      rq1_lrt_chi2     = if (!is.null(lrt)) lrt$Chisq[2] else NA_real_,
      rq1_lrt_p        = if (!is.null(lrt)) lrt$`Pr(>Chisq)`[2] else NA_real_,
      rq1_aic_full     = if (!is.null(rq1_full)) AIC(rq1_full) else NA_real_,
      rq1_aic_reduced  = if (!is.null(rq1_reduced)) AIC(rq1_reduced) else NA_real_,
      rq1_aic_delta    = rq1_aic_reduced - rq1_aic_full
    )
  }) %>%
  bind_rows() %>%
  arrange(n_params)



```

### Final step

First, we report results for the final step.

```{r tables}
developmental_results_rq1 %>%
  filter(checkpoint == 143000) %>%
  mutate(
    rq1_lrt_p = format.pval(rq1_lrt_p, digits = 3),
    across(c(rq1_dom_beta, rq1_dom_se, rq1_lrt_chi2, rq1_aic_delta), ~ round(.x, 3))
  ) %>%
  select(model, n_params, rq1_dom_beta, rq1_dom_se, rq1_lrt_chi2, rq1_lrt_p, rq1_aic_delta) %>%
  kable(
    col.names = c("Model", "Params", "β (Dominance)", "SE", "χ²", "p", "ΔAIC"),
    caption = "RQ1: Dominance predicting target surprisal (final checkpoint)"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

developmental_results_rq1 %>%
  filter(checkpoint == 143000) %>%
  mutate(
    rq1_lrt_p = format.pval(rq1_lrt_p, digits = 3),
    across(c(rq1_dom_beta, rq1_dom_se, rq1_lrt_chi2, rq1_aic_delta), ~ round(.x, 3))
  ) %>%
  select(model, n_params, rq1_dom_beta, rq1_dom_se, rq1_lrt_chi2, rq1_lrt_p, rq1_aic_delta) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    col.names = c("Model", "Params", "$\\beta$", "SE", "$\\chi^2$", "$p$", "$\\Delta$AIC"),
    escape = FALSE,
    caption = "RQ1: Dominance predicting target surprisal (final checkpoint)"
  ) %>%
  kable_styling(latex_options = c("hold_position"))
```

We also visualize the coefficients.

```{r coef_rq1_final}
developmental_results_rq1 %>%
  filter(checkpoint == 143000) %>%
  mutate(
    model_shorthand = sub("^EleutherAI/", "", model),
    model_shorthand = fct_reorder(model_shorthand, n_params)
  ) %>%
  ggplot(aes(x = model_shorthand, y = rq1_dom_beta)) +
  geom_pointrange(aes(
    ymin = rq1_dom_beta - 1.96 * rq1_dom_se,
    ymax = rq1_dom_beta + 1.96 * rq1_dom_se
  )) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    x = "Model (ordered by parameter count)",
    y = "β (Dominance)",
    title = "RQ1: Effect of sense dominance on target surprisal",
    subtitle = "Point estimates ± 95% CI"
  ) +
  coord_flip() +
  theme_minimal(base_size = 13) 
```


And also the raw data.

```{r rq1_raw_data_final}
df_merged %>%
  filter(checkpoint == 143000) %>%
  ggplot(aes(x = dominance_right,
             y = target_surprisal_mean)) +
  geom_point(alpha = .5) +
  geom_smooth(method = "lm") +
  labs(x = "Human Dominance Rating",
       y = "Surprisal on Target") +
  theme_minimal(base_size = 13) +
  facet_wrap(~reorder(model_shorthand, n_params))

```


### RQ5: Model scaling

```{r rq1_scaling}

## scaling_results_rq1
developmental_results_rq1 %>%
  filter(checkpoint == 143000) %>%
  ggplot(aes(x = n_params,
             y = rq1_aic_delta)) +
  geom_point(size = 6,
             alpha = .5) +
  scale_x_log10() +
  geom_text_repel(aes(label=model_shorthand), size=3) +
  labs(x = "Parameters",
       y = "AIC Delta") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") 

developmental_results_rq1 %>%
  filter(checkpoint == 143000) %>%
  ggplot(aes(x = n_params,
             y = rq1_dom_beta)) +
  geom_point(size = 6,
             alpha = .5) +
  scale_x_log10() +
  geom_text_repel(aes(label=model_shorthand), size=3) +
  labs(x = "Parameters",
       y = "Dominance Coefficient") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") 



summary(lm(rq1_dom_beta ~ log10(n_params),
           data = developmental_results_rq1 %>%
  filter(checkpoint == 143000)))

summary(lm(rq1_aic_delta ~ log10(n_params),
           data = developmental_results_rq1 %>%
  filter(checkpoint == 143000)))

```


### RQ6: Developmental Trajectory

```{r rq6_dev}

developmental_results_rq1 %>%
  mutate(significant = rq1_lrt_p < .05) %>%
  ggplot(aes(x = checkpoint,
             y = rq1_dom_beta,
             color = reorder(model_shorthand, n_params))) +
  geom_point(aes(shape = significant),
             size = 4, alpha = .7) +
  geom_line(size = 1.3) +
  geom_point(size = 4, alpha = .4) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Coefficient (Dominance)",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  scale_shape_manual(values = c("FALSE" = 1, "TRUE" = 16),
                     labels = c("n.s.", "p < .05"),
                     name = NULL) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(7, option = "magma", 
                                                begin = 0.8, end = 0.15))
```

## RQ3: Does surprisal track sense boundaries?

```{r}
developmental_results_rq3 <- df_pythia_models %>%
  mutate(
    ambiguity_c = if_else(ambiguity_type == "Polysemy", 0.5, -0.5),
    same_c      = if_else(same == TRUE, 0.5, -0.5)
  ) %>%
  group_by(model, model_shorthand, n_params, checkpoint) %>%
  group_map(~ {
    rq3_full    <- tryCatch(lmer(target_surprisal_mean ~ same_c + ambiguity_c + (1 | word), data = .x, REML = FALSE), error = function(e) NULL)
    rq3_reduced <- tryCatch(lmer(target_surprisal_mean ~           ambiguity_c + (1 | word), data = .x, REML = FALSE), error = function(e) NULL)
    lrt <- if (!is.null(rq3_full) && !is.null(rq3_reduced)) anova(rq3_full, rq3_reduced) else NULL
    tibble(
      model            = .y$model,
      model_shorthand  = .y$model_shorthand,
      n_params         = .y$n_params,
      checkpoint       = .y$checkpoint,
      rq3_same_beta    = if (!is.null(rq3_full)) fixef(rq3_full)["same_c"] else NA_real_,
      rq3_same_se      = if (!is.null(rq3_full)) sqrt(diag(vcov(rq3_full)))["same_c"] else NA_real_,
      rq3_lrt_chi2     = if (!is.null(lrt)) lrt$Chisq[2] else NA_real_,
      rq3_lrt_p        = if (!is.null(lrt)) lrt$`Pr(>Chisq)`[2] else NA_real_,
      rq3_aic_full     = if (!is.null(rq3_full)) AIC(rq3_full) else NA_real_,
      rq3_aic_reduced  = if (!is.null(rq3_reduced)) AIC(rq3_reduced) else NA_real_,
      rq3_aic_delta    = rq3_aic_reduced - rq3_aic_full
    )
  }) %>%
  bind_rows() %>%
  arrange(n_params)

```


### Final step

First, we report results for the final step.

```{r tables_rq3}
developmental_results_rq3 %>%
  filter(checkpoint == 143000) %>%
  mutate(
    rq3_lrt_p = format.pval(rq3_lrt_p, digits = 3),
    across(c(rq3_same_beta, rq3_same_se, 
             rq3_lrt_chi2, rq3_aic_delta), ~ round(.x, 3))
  ) %>%
  select(model, n_params, rq3_same_beta, rq3_same_se, rq3_lrt_chi2, rq3_lrt_p, rq3_aic_delta) %>%
  kable(
    col.names = c("Model", "Params", "β (Same Sense)", "SE", "χ²", "p", "ΔAIC"),
    caption = "RQ3: Sense Boundary predicting target surprisal (final checkpoint)"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

developmental_results_rq3 %>%
  filter(checkpoint == 143000) %>%
  mutate(
    rq3_lrt_p = format.pval(rq3_lrt_p, digits = 3),
    across(c(rq3_same_beta, rq3_same_se, 
             rq3_lrt_chi2, rq3_aic_delta), ~ round(.x, 3))
  ) %>%
  select(model, n_params, rq3_same_beta, rq3_same_se, rq3_lrt_chi2, rq3_lrt_p, rq3_aic_delta) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    col.names = c("Model", "Params", "$\\beta$", "SE", "$\\chi^2$", "$p$", "$\\Delta$AIC"),
    escape = FALSE,
    caption = "RQ3: Sense Boundary predicting target surprisal (final checkpoint)"
  ) %>%
  kable_styling(latex_options = c("hold_position"))
```

We also visualize the coefficients.

```{r coef_rq3_final}
developmental_results_rq3 %>%
  filter(checkpoint == 143000) %>%
  mutate(
    model_shorthand = sub("^EleutherAI/", "", model),
    model_shorthand = fct_reorder(model_shorthand, n_params)
  ) %>%
  ggplot(aes(x = model_shorthand, y = rq3_same_beta)) +
  geom_pointrange(aes(
    ymin = rq3_same_beta - 1.96 * rq3_same_se,
    ymax = rq3_same_beta + 1.96 * rq3_same_se
  )) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    x = "Model (ordered by parameter count)",
    y = "β (Sense Boundary)",
    title = "RQ1: Effect of sense boundary on target surprisal",
    subtitle = "Point estimates ± 95% CI"
  ) +
  coord_flip() +
  theme_minimal(base_size = 13) 
```


And also the raw data.

```{r rq3_raw_data_final}
df_pythia_models %>%
  filter(checkpoint == 143000) %>%
  ggplot(aes(x = target_surprisal_mean,
             fill = same_label,
             y = ambiguity_type)) +
  geom_density_ridges(color = NA, alpha = 0.7, scale = 0.9) +
  scale_fill_manual(values = c("Same Sense" = "#4E79A7", "Different Sense" = "#E15759")) +
  labs(x = "Mean target surprisal",
       y = NULL,
       fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    legend.position = "top"
  ) +
  facet_wrap(~reorder(model_shorthand, n_params))


df_pythia_models %>%
  filter(model_shorthand == "pythia-2.8b") %>%
  ggplot(aes(x = target_surprisal_mean,
             fill = same_label,
             y = ambiguity_type)) +
  geom_density_ridges(color = NA, alpha = 0.7, scale = 0.9) +
  scale_fill_manual(values = c("Same Sense" = "#4E79A7", "Different Sense" = "#E15759")) +
  labs(x = "Mean target surprisal",
       y = NULL,
       fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    legend.position = "top"
  ) +
  facet_wrap(~checkpoint)

```

### RQ5: Model scaling

```{r rq3_scaling}

## scaling_results_rq1
developmental_results_rq3 %>%
  filter(checkpoint == 143000) %>%
  ggplot(aes(x = n_params,
             y = rq3_aic_delta)) +
  geom_point(size = 6,
             alpha = .5) +
  scale_x_log10() +
  geom_text_repel(aes(label=model_shorthand), size=3) +
  labs(x = "Parameters",
       y = "AIC Delta") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") 

developmental_results_rq3 %>%
  filter(checkpoint == 143000) %>%
  ggplot(aes(x = n_params,
             y = rq3_same_beta)) +
  geom_point(size = 6,
             alpha = .5) +
  scale_x_log10() +
  geom_text_repel(aes(label=model_shorthand), size=3) +
  labs(x = "Parameters",
       y = "Sense Boundary Coefficient") +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") 



summary(lm(rq3_same_beta ~ log10(n_params),
           data = developmental_results_rq3 %>%
  filter(checkpoint == 143000)))

summary(lm(rq3_aic_delta ~ log10(n_params),
           data = developmental_results_rq3 %>%
  filter(checkpoint == 143000)))

```


### RQ6: Developmental Trajectory

```{r rq6_dev_senses}

developmental_results_rq3 %>%
  mutate(significant = rq3_lrt_p < .05) %>%
  ggplot(aes(x = checkpoint,
             y = rq3_same_beta,
             color = reorder(model_shorthand, n_params))) +
  geom_point(aes(shape = significant),
             size = 4, alpha = .7) +
  geom_line(size = 1.3) +
  geom_point(size = 4, alpha = .4) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Coefficient (Sense Boundary)",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, 
             linetype = "dotted", 
             size = 1.2) +
  scale_shape_manual(values = c("FALSE" = 1, "TRUE" = 16),
                     labels = c("n.s.", "p < .05"),
                     name = NULL) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(7, option = "magma", 
                                                begin = 0.8, end = 0.15))
```


